Analysis of transit delays in the Toronto Transit Commission: 2024

### 2024-09-25 paper/paper.qmd
Update paper.qmd

# Introduction
Having operated for over 100 years, the Toronto Transit Commission (TTC) has become one of the most visible and representative public transportation organizations in North America. It is well-known as the primary mode of public transportation in Toronto's daily life. Within the TTC, the bus network typically experiences higher daily ridership compared to streetcars, while subways accommodate the largest volume of passengers due to their central role in covering longer routes. Frequent riders often notice delays that arise on a daily basis. As delays become worse and services becoming increasingly inconsistent, it is important to examine the potential causes, allowing the TTC to identify and implement effective solutions.

@@

# Data {#sec-data}
## Data Source
The data are gathered from Open Data Toronto's report on TTC delays. The datasets, published by the TTC, include: the first dataset, which presents delay statistics for subways in 2024; the second dataset, which details delay statistics for buses in 2024; and the third dataset, which provides delay statistics for streetcars in 2024.

@@

>>>>

# Introduction

Having operated for over 100 years, the Toronto Transit Commission (TTC) has become one of the most visible and representative public transportation organizations in North America. It is well-known as the primary mode of public transportation in Toronto's daily life. Within the TTC, the bus network typically experiences higher daily ridership compared to streetcars, while subways accommodate the largest volume of passengers due to their central role in covering longer routes. Frequent riders often notice delays that arise on a daily basis. As delays become worse and services becoming increasingly inconsistent, it is important to examine the potential causes, allowing the TTC to identify and implement effective solutions.

@@

# Data {#sec-data}
## Data Source
The data were gathered from Open Data Toronto's report on TTC delays. The datasets, published by the TTC, include the following: the first dataset presents delay statistics for subways in 2024 from January 1 to August 31; the second dataset details delay statistics for buses in 2024 from January 1 to August 31; and the third dataset provides delay statistics for streetcars in 2024 from January 1 to August 31.

@@

# Overview of datasets
```{r}
library(dplyr)
library(knitr)
data <- read.csv("ttc_bus_delay_data_2024.csv")

subset_data <- data %>%
  select(Date, Time, Day, Incident, Min.Delay, Min.Gap) %>% 
  head(10)

# Print the data with lines between columns
kable(subset_data) 
```
Table 1: Sample of bus delay statistics; the first 10 rows of TTC bus delay over time: Jan 1 to Aug 31, 2024.

```{r}
library(dplyr)
library(knitr)
data <- read.csv("ttc_subway_delay_data_2024.csv")

subset_data <- data %>%
  select(Date, Time, Day, Code, Min.Delay, Min.Gap) %>% 
  head(10)

# Print the data with lines between columns
kable(subset_data) 
```
Table 2: Sample of subway delay statistics; the first 10 rows of TTC subway delay over time: Jan 1 to Aug 31, 2024.

Both Table 1 and Table 2 include:

1. Date: The dataset shows the date when the delay occurred.
2. Time: The dataset contains information on the time the delay occurred.
3. Day: The dataset indicates the day of the week when the delay occurred.
4. Incident: The dataset displays the reason for the delay, such as a general delay, emergency services, or security issues.
5. Min.Delay and Min.Gap: The dataset likely represents 'Min Delay' and 'Min Gap,' both measured in minutes. 'Min Delay' refers to the delay in the schedule for the next bus, while 'Min Gap' indicates the scheduled time, between the bus ahead and the following bus.

The total bus delay dataset contains 40,447 entries, while the subway delay dataset contains 17,517 entries. We can indicate that delays in the bus has occurred approximately 2.35 times more than the subway delay.

Both datasets went through basic cleaning for readability and irrelevant or unusable columns were removed and not considered in this report.
Originally, the subway data reported the incident causing the delay as a unique code rather than an understandable description. This description was included instead in a separate table from the same package (Data 2023b). Using merge from dplyr (Wickham et al. 2023), the incident was mapped to a generalized version of the description to match with the same set of generalized descriptions in the bus delay data. This selection into sections based on the description important to note due to the potential subjectivity of the sorting. There are incident descriptions in which the record may fall in one or the other. Please see the data cleaning scripts for more details.

Relate your analysis to the themes mentioned in Bouie (2022), discussing the importance of remembering and analyzing historical events through data.
